Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-07-12 manuscript/manuscript.tex
Continue writing analytical solutions

\documentclass[extra]{gji}
%~ %\documentclass[extra, referee]{gji}

@@

@@

\noindent $\alpha, \beta \in \{x, y, z\}$,
$\rho(r')$ is the density function that depends on the radius coordinate,
$\delta_{\alpha\beta}$ is Kronecker's delta,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant and
\begin{equation}
    \Delta_x = r'[\cos\phi\sin\phi' - \sin\phi\cos\phi'
               \cos(\lambda' - \lambda)],
\end{equation}
\begin{equation}
    \Delta_y = r' \cos \phi' \sin(\lambda' - \lambda),
\end{equation}
\begin{equation}
    \Delta_z = r' \cos \psi - r,
\end{equation}
\begin{equation}
    \kappa = {r'}^2 \cos \phi',
\end{equation}
\begin{equation}
    \ell = \sqrt{{r'}^2 + r^2 - 2 r' r \cos \psi},
\end{equation}
\begin{equation}
    \cos\psi = \sin\phi\sin\phi' + \cos\phi\cos\phi'
                 \cos(\lambda' - \lambda).
\end{equation}

@@

>>>>

\documentclass[extra]{gji}
%~ \documentclass[extra, referee]{gji}

@@

\todo{aclarar coordenadas esfericas -> longitud latitud angulos}

@@

\noindent $\alpha, \beta \in \{x, y, z\}$,
$\rho(r')$ is the density function that depends on the radius coordinate,
$\delta_{\alpha\beta}$ is Kronecker's delta,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant and
\begin{equation}
    \Delta_x = r'[\cos\phi\sin\phi' - \sin\phi\cos\phi'
               \cos(\lambda' - \lambda)],
\end{equation}
\begin{equation}
    \Delta_y = r' \cos \phi' \sin(\lambda' - \lambda),
\end{equation}
\begin{equation}
    \Delta_z = r' \cos \psi - r,
\end{equation}
\begin{equation}
    \kappa = {r'}^2 \cos \phi',
\end{equation}
\begin{equation}
    \ell = \sqrt{{r'}^2 + r^2 - 2 r' r \cos \psi},
\label{eq:ell}
\end{equation}
\begin{equation}
    \cos\psi = \sin\phi\sin\phi' + \cos\phi\cos\phi'
                 \cos(\lambda' - \lambda).
\label{eq:cospsi}
\end{equation}

@@

\subsection{Analytical Solutions for Spherical Shell}

In order to test the accuracy of the GLQ approximation, \citet{Uieda2016} and \citet{Grombein2013} have compared the computed gravity fields with the ones generated by a spherical shell, the exceptional case with analytical solution \citep{Mikuska2006,Grombein2013}.

In case of variable density tesseroids, the comparison must obviously be made with a variable density spherical shell, whose analytical solution will depend on the chosen density function.
Because the literature has not offered it yet, we needed to develop ourselves.

Lets consider a spherical shell with inner and outer radii $R_1$ and $R_2$, respectivetly, which density depends on the radius spherical coordinate.
The gravity potential generated by the shell on an arbitrary external point $Q(0,0,r)$ located along the $z$ axis at a distance $r$ can be written as follows:

\todo{figura spherical shell con punto Q}

\begin{equation}
    V_\text{sh}(r) = G 
    \int\limits_0^{2\pi}
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r')}{\ell} {r'}^2 \cos\phi' \, 
    dr' d\phi' d\lambda',
\end{equation}

\noindent where $\ell$ is defined as in equation \ref{eq:ell}.
The computation point $Q$ is located at latitude $\phi=90^\circ$, so $\ell$ and $\cos\psi$ (defined in equation \ref{eq:cospsi}) get simplified:

\begin{equation}
    \cos\psi = \sin\phi', \quad
    \ell = \sqrt{r'^2 + 2 r' r \sin\phi' + r^2}.
\end{equation}

Due to the rotational symmetry along the $z$ axis, the integration in $\lambda'$ is straightforward:

\begin{equation}
    V_\text{sh}(r) = 2\pi G 
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r') {r'}^2 \cos\phi'}{\sqrt{r'^2 + 2 r' r \sin\phi' + r^2}}
    \, dr' d\phi',
\end{equation}

\noindent and the integration in $\phi'$ can be performed independently of the density function.
Using \citet{sagemath} we obtain the following expression:

\iftwocol{
\begin{equation}
\begin{split}
    V_\text{sh}(r) = 2\pi G
    \int\limits_{R_1}^{R_2}
    \Big[ \sqrt{r^2 + r'^2 + 2rr'} -& \\
    \sqrt{r^2 + r'^2 - 2rr'}
    & \Big] \frac{r'\rho(r')}{r} \, dr'.
\end{split}
\label{eq:shell-pot-sqrts}
\end{equation}
}
{
\begin{equation}
    V_\text{sh}(r) = 2\pi G
    \int\limits_{R_1}^{R_2}
    \Big[ \sqrt{r^2 + r'^2 + 2rr'}  -
    \sqrt{r^2 + r'^2 - 2rr'} 
    \Big] \frac{r'\rho(r')}{r} \, dr'.
\label{eq:shell-pot-sqrts}
\end{equation}
}

For our purposes we can assume that the computation point $Q$ is outside the outer radii, i.e. $r>R_2$, and simplify the square roots in equation \ref{eq:shell-pot-sqrts}:

\begin{equation}
    \sqrt{r^2 + r'^2 + 2rr'} = |r + r'| = r + r',
\end{equation}
\begin{equation}
    \sqrt{r^2 + r'^2 - 2rr'} = |r - r'| = r - r',
\end{equation}

\noindent what leads to the following expression of the potential of the spherical shell:

\begin{equation}
    V_\text{sh}(r) = \frac{4\pi G}{r}
    \int\limits_{R_1}^{R_2} {r'}^2 \rho(r') \, dr'.
\label{eq:shell-pot}
\end{equation}

The equation \ref{eq:shell-pot} allows to easily obtain the exact gravity field generated by a spherical shell with a variable density in depth on any outside point located in the $z$ axis.
Due to the rotational symmetry along any axe that passes through the centre of the shell, the potential in equation \ref{eq:shell-pot} reassembles the gravity field on any outside point at distance $r$ from the its centre.
